National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs

### 2020-11-19 Manuscript.tex
Updates from Overleaf

\title{National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs}
\author{Ben Highland\textsuperscript{1}, Everett L. Worthington\textsuperscript{2}, Don E Davis\textsuperscript{3}, Chris G. Sibley\textsuperscript{4}, \& Joseph Bulbulia\textsuperscript{5}}
\date{}

@@

1 Wake Forest Medical School 2 Department of Psychology Virginia Commonwealth University 3 Department of Psychology Georgia State University 4 School of Psychology University of Auckland 5 School of Psychology Victoria University

Correspondence concerning this article should be addressed to Ben Highland, Ben's Postal address. E-mail: \href{mailto:benhighland6420@gmail.com}{\nolinkurl{benhighland6420@gmail.com}}

@@

\abstract{
Previous research finds an association between spirituality and subjective well-being. However, the widespread use of tautological spirituality scales, poorly defined concepts of spirituality, and heavy reliance on cross-sectional samples cast doubts. Here, we leverage nine waves of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction (\(N\) = 20979, New Zealand, 2010-2018). Contrary to previous research, we find that belief in a spirit or life force predicts lower personal well-being and life-satisfaction. However, in support of previous speculation, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief. These finding are robust after to known demographic influences, and intriguingly, hold among those who believe in a God but disbelieve in a Spirit or Life Force. The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates further causal investigations of the mechanisms by which spiritual beliefs lead to growth in subjective well-being.
}

@@

First, previous research has frequently operationalised spirituality using measures of well-being. For example, The \emph{Daily Spiritual Experience Scale}, which is widely used in spiritual mental health research, includes items such as \enquote{I feel thankful for my blessings} and \enquote{I feel deep inner peace or harmony} to measure spirituality (Underwood \& Teresi, 2002). As Koenig observes, it is unsurprising that spirituality predict well-being when the same items are used to measure both constructs. Koenig aptly describes this approach as \enquote{tautological} (Koenig, 2008). Indeed, recent systematic-review of spirituality/mental health research found that nearly 45\% of previous studies employe tautological scales (Garssen et al., 2016). The first challenge, then, is to define measures of spirituality that are not well-being measures dressed up in other words.

A second challenge is to disentangle concepts of spirituality from concepts of religion. The term \enquote{spirituality} was first used to describe the religious practices of ascetics and monks, however, more recently, the term has taken on a broader family of meanings (Swinton, 2001). Spirituality and health research does not converge on a unique set of meanings from the wider universe of popular meanings. Although it is commonplace across most areas of psychological science for researchers to explore different operationalisations of focal concepts, the psychology of spirituality has mostly been conducted with cross-sectional North American samples among participants who affiliate with traditional religion (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky, Ratner, \& Chiu, 2005; Smith et al., 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012). As such, the extent to which the psychology of spirituality and health has repeated or extended the psychology of religion and health remains unclear. Notably, researchers who have operationalised spirituality constructs as distinct from religion constructs have observed a negative association between spirituality and subjective well-being (King et al., 2013). Thus to infer how spirituality affects subjective well-being, it is important to disentangle the concepts of spirituality from concepts of religion, and to carefully investigate features of contructs in populations where these dimensions overlap, partially overlap, and diverge.

A third challenge is to investigate the relationship between spirituality and subjective well-being in nationally diverse samples over individual life-spans. Reliable inference requires observing change within people who differ dimensions of spirituality over time, and assing how their lives turn out (Garssen \& Visser, 2016). Previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have relied on cross-sectional samples (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky et al., 2005; Smith et al., 2003; Visser et al., 2010; Yonker et al., 2012). However, To address the key questions in spirituality psychological health research requires longitudinal investigations. Unfortunately the relevant datasets are rare.

@@

Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05). We analyze data from participants who responded to the NZAVS survey at least three four times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of 21,705 (N = 20,979 with complete responses). To model growth in subjective well-being we use generalized linear mixed models. In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator). The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God. To handle the dependencies and heterogeneity introduced from the repeated measures, we include individual ID as an effect modelled as random. We centered and scaled the education, deprivation, and political conservativism. We centred age was centred at and put into decade-long intervals.
The model equation(s) are described in Equation 1.

@@

Because missing values can result in biased estimates (Blackwell, Honaker, \& King, 2017), we multiply imputed 10 datasets using the AmeliaII package in R(Honaker et al., 2011b).We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences. Details of this analysis can be found at: \url{https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing}.

@@

\begin{itemize}
\tightlist
\item
  The effect of Years is positive is not reliable (beta = 2.66e-03, SE = 3.58e-03, std. beta = 3.19e-03, p = 0.458). We infer that Skeptics personal well-being is not increasing in the Skeptic population
\item
  The effect of Beliefs in SpiritExcludesGod is reliably negative (beta = -0.14, SE = 0.03, std. beta = -0.03, p \textless{} .001). We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower personal well-being than skeptics.
\item
  The effect of Beliefs in GodAndSpirit is reliably negative (beta = -0.16, SE = 0.03, std. beta = -0.02, p \textless{} .001). We infer that overall the population that believes in God and a spirit or life force is expected to have a lower Personal Well-being than skeptics.
\item
  The effect of Beliefs in GodExcludesSpirit is not reliably different from the skeptic population (beta = 0.02, SE = 0.05, std. beta = 0.02, p = 0.669).
\item
  The effect of Age.10yrs.C is reliably positive (beta = 0.08, SE = 7.38e-03, std. beta = 0.07, p \textless{} .001). Older people have greater personal well-beingwellbeing.
\item
  The effect of Deprivation (S) is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p \textless{} .001). Deprived people have lower personal well-beingwellbeing.
\item
  The effect of Education (S) is reliably positive (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p \textless{} .001). Educated people greater personal well-beingwellbeing
\item
  The effect of Employed is reliably positive (beta = 0.09, SE = 0.01, std. beta = 0.05, p \textless{} .001). Employed people have greater personal well-beingwellbeing.
\item
  The effect of identifying as Maori is reliably negative (beta = -0.30, SE = 0.03, std. beta = -0.18, p \textless{} .001).
\item
  The effect of identifying as Pacific is reliably negative (beta = -0.29, SE = 0.05, std. beta = -0.17, p \textless{} .001).
\item
  The effect of identifying as Asian is not reliable (beta = -0.08, SE = 0.05, std. beta = -0.05, p = 0.089).
\item
  The effect of identifying as Male is reliably negative (beta = -0.11, SE = 0.02, std. beta = -0.06, p \textless{} .001).
\item
  The effect of having a Partner is reliably positive (beta = 0.61, SE = 0.02, std. beta = 0.37, p \textless{} .001).
\item
  The effect of Political Conservativism (S) is reliably positive (beta = 0.02, SE = 5.67e-03, std. beta = 0.01, p \textless{} .001).
\item
  The effect of living in an Urban area is reliably negative (beta = -0.02, SE = 0.01, std. beta = -0.01, p \textless{} .05).
\end{itemize}

\hypertarget{personal-well-being-years-x-beliefs}{%
\subsection{Personal well-being: years X beliefs}\label{personal-well-being-years-x-beliefs}}

@@

\begin{itemize}
\tightlist
\item
  The effect of years \(\times\) belief in a SpiritExcludesGod\_ is reliably positive (beta = 0.02, SE = 4.86e-03, std. beta = 0.02, p \textless{} .01). We infer that, unlike the skeptic population, people with beliefs in a spirit or life force who do not believe in a God tend to grow in their personal well-being over time.
\item
  The effect of years \(\times\) in GodAndSpirit is reliably positive (beta = 0.02, SE = 4.41e-03, std. beta = 0.03, p \textless{} .001). We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their personal well-being over time
\item
  The effect of Years \(\times\) belief in GodExcludesSpirit is not reliable(beta = 1.96e-03, SE = 8.29e-03, std. beta = 2.36e-03, p = 0.813). Thus, God believers who doubt in a Spirit or Life Force start maintain higher personal well-being than do other groups, but they are not expected to grow in personal well-being over time.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  The effect of years is not reliable (beta = -5.50e-04, SE = 2.69e-03, std. beta = -9.32e-04, p = 0.838). We do not infer growth over time in Life Satisfaction among the Skeptic population
\item
  The effect of beliefs in SpiritExcludesGod is reliably negative (beta = -0.08, SE = 0.02, std. beta = 0.01, p \textless{} .001). We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower Life Satisfaction than skeptics.
\item
  The effect of beliefs in GodAndSpirit is not reliable (beta = -0.04, SE = 0.02, std. beta = 0.04, p = 0.105). We do not infer growth over time in Life Satisfaction among this (Skeptic) population
\item
  The effect of beliefs in GodExcludesSpirit is reliably positive (beta = 0.07, SE = 0.04, std. beta = 0.05, p = 0.062).We infer that overall the population that believes in a God but does not believe in a spirit is not reliably different from the Skeptics population in Life Satisfaction.
\item
  The effect of Age.10yrs.C is reliably positive (beta = 0.09, SE = 5.26e-03, std. beta = 0.10, p \textless{} .001). Older people have greater Life Satisfaction
\item
  The effect of Deprivation (S) is is reliably negative (beta = -0.07, SE = 5.50e-03, std. beta = -0.06, p \textless{} .001). People who are more deprived have lower Life Satisfaction.
\item
  The effect of Education (S) is reliably positive (beta = 0.06, SE = 6.33e-03, std. beta = 0.05, p \textless{} .001). Educated people have greater Life Satisfaction
\item
  The effect of being Employed is reliably positive (beta = 0.08, SE = 9.69e-03, std. beta = 0.07, p \textless{} .001). People who are employed tend to have greater Life Satisfaction
\item
  The effect of identifying as Maori is reliably negative (beta = -0.06, SE = 0.02, std. beta = -0.05, p \textless{} .01). People who belong to the minority Maori group tend to have lower Life Satisfaction.
\item
  The effect of identifying as Pacific is reliably negative (beta = -0.08, SE = 0.04, std. beta = -0.07, p \textless{} .05). People who belong to the minority Pacific peoples group tend to have lower life satisfaction.
\item
  The effect of identifying as Asian is not reliably different from people of European ethnicity in life satisfaction(beta = -0.04, SE = 0.03, std. beta = -0.03, p = 0.223).
\item
  The effect of identifying as Male is reliably negative (beta = -0.19, SE = 0.01, std. beta = -0.16, p \textless{} .001).
\item
  The effect of having a Parnter is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p \textless{} .001).
\item
  The effect of Political Conservativism (S) is reliably positive (beta = 0.01, SE = 4.23e-03, std. beta = 0.01, p \textless{} .01).
\item
  The effect of living in an Urban area is reliably negative (beta = -0.02, SE = 8.85e-03, std. beta = -0.02, p \textless{} .05).
\end{itemize}

\hypertarget{life-satisfaction-years-x-beliefs}{%
\subsection{Life Satisfaction: years X beliefs}\label{life-satisfaction-years-x-beliefs}}

@@

\begin{itemize}
\tightlist
\item
  The interaction of years \(\times\) beliefs in SpiritExcludesGod is reliably positive (beta = 0.02, SE = 3.66e-03, std. beta = 0.03, p \textless{} .001). We infer that, unlike the Skeptic population, people with beliefs in a Spirit or Life Force who do not believe in a God tend to grow in their Life Satisfaction over time.

  \begin{itemize}
  \tightlist
  \item
    The interaction of years \(\times\) beliefsGodAndSpirit is reliably positive (beta = 0.01, SE = 3.32e-03, std. beta = 0.02, p \textless{} .001). We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their Life Satisfaction over time
  \item
    The effect of years \(\times\) beliefs in GodExcludesSpirit is not reliable (beta = -1.59e-03, SE = 6.25e-03, std. beta = -2.70e-03, p = 0.799). Thus, God believers who doubt in a Spirit or Life Force start maintain higher Life-Satisfaction than do other groups, but they are not expected to grow in Life-Satisfaction over time.
  \end{itemize}
\end{itemize}

@@

To adjust for potential bias from missing responses, we conducted a supplementary analysis in which we imputed missing values over the the longitudinal sample. This analysis is describe in Supplement 1. Averaging over the missing responses does not change inference, however we observe growth across all four groups in both personal well-being and life satisfaction. Additionally we observe substantially stronger effects for the slope of Spirit Beliefs on both personal well-being and life-satisfaction. Figures 3 and Figures 4 describe the predicted slopes of change. The inference from this analysis remains the same as from the analysis in which we delete missing values row-wise, by panel wave. In all analyses, those who believe in a spirit or life-force experience greater growth in both Personal Well-being and Life Satisfaction over time.

@@

  At the outset of the study, in 2010, the expected subjective well-being as measured using the Personal Well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God (\(\hat{y}\) = 6.50, se =.03) and lower among those who believed in both a Spirit/Life-Force and a God (\(\hat{y}\) = 6.48, se =.03) in comparison both to the skeptical population who neither believed in a God nor a life force (\(\hat{y}\) = 6.64,se =.03) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 6.66, se =.05). (See Figure 2) Similarly,the expected subjective well-being as measured using the Life Satisfaction Inventory is lower among those who believed in Spirit or Life Force but not a God (\(\hat{y}\) = 4.86, se =.02) and lower among those who believed in both a Spirit/Life-Force and a God (\(\hat{y}\) = 4.91, se =.02) in comparison both with the skeptical population who neither believed in a God nor a life force (\(\hat{y}\) = 4.94,se =.02) and the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 5.02, se =.02) (see Figure 3)

@@

  However, over time, belief in a Spirit or Life Force predicts a trajectory of growth in personal well-being. In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to (\(\hat{y}\) = 6.67, se =.03, \(\Delta \hat{y} = 0.17\)). Among those who believed in both a Spirit/Life-Force and a God Personal Well-Being had grown to (\(\hat{y}\) = 6.69, se = 0.02,\(\Delta \hat{y} = .21\)). To place this effect in practical perspective, such growth in personal well-being is about twice as large as the expected benefit of being employed and about a third as large as the expected benefit from having a romantic partner. By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force remained stable across the eight year interval \(\hat{y} = 6.66, se = 0.03, \Delta \hat{y} = .02\)). By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force was constant \(\hat{y} = 6.70, se = 0.04, \Delta \hat{y} = .04\)). Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in Life Satisfaction. In 2018, life satisfaction among those who believe in Spirit or Life Force but not a God had grown to (\(\hat{y} = 5.01, se = 0.02, \Delta \hat{y} = 0.07\)). In practical terms, such growth is equivalent to the expected gain in Life Satisfaction from being employed and a sixth as large as the expected benefit to Life Satisfaction from having a romantic partner. Among those who believed in both a Spirit/Life-Force and a God, life satisfaction had grown to (\(\hat{y} = 5.03, se = 0.02, \Delta \hat{y} = 0.12\)). In practical terms, such growth is equivalent to 1.5 times the expected gain to life satisfaction from being employed and almost a third benefit to life satisfaction from having a romantic partner. By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force remained stable over the eight year interval \(\hat{y} = 4.94, se =0.02, \Delta \hat{y} = 0\)). Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force was constant \(\hat{y} = 5.00, se = 0.03, \Delta \hat{y} = -.02\)) (See Figure 3).

>>>>

\title{National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs}
\author{Benjamin Highland\textsuperscript{1}, Everett L. Worthington\textsuperscript{2}, Don E Davis\textsuperscript{3}, Chris G. Sibley\textsuperscript{4}, \& Joseph Bulbulia\textsuperscript{5}}
\date{}

@@

1 Wake Forest School of Medicine 2 Department of Psychology Virginia Commonwealth University 3 Department of Psychology Georgia State University 4 School of Psychology University of Auckland 5 School of Psychology Victoria University

Correspondence concerning this article should be addressed to Benjamin Highland, Benjamin's E-mail: \href{mailto:benhighland6420@gmail.com}{\nolinkurl{benhighland6420@gmail.com}}

@@

\abstract{Previous research finds an association between spirituality and subjective well-being. However, the widespread use of tautological spirituality scales, poorly defined concepts of spirituality, and heavy reliance on cross-sectional samples cast doubts. Here, we leverage nine waves of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction (\(N\) = 20979, New Zealand, 2010-2018). Contrary to previous research, we find that belief in a spirit or life force predicts lower personal well-being and life-satisfaction. However, in support of previous speculation, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief. These finding are robust after adjusting for known demographic influences, and intriguingly, hold among those who believe in a God but disbelieve in a Spirit or Life Force. The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates further causal investigations of the mechanisms by which spiritual beliefs lead to growth in subjective well-being.
}

@@

First, previous research has frequently operationalised spirituality using measures of well-being. For example, The \emph{Daily Spiritual Experience Scale}, which is widely used in spiritual mental health research, includes items such as \enquote{I feel thankful for my blessings} and \enquote{I feel deep inner peace or harmony} to measure spirituality (Underwood \& Teresi, 2002). As Koenig observes, it is unsurprising that spirituality predict well-being when the same items are used to measure both constructs. Koenig aptly describes this approach as \enquote{tautological} (Koenig, 2008). Indeed, recent systematic-review of spirituality/mental health research found that nearly 45\% of previous studies employ tautological scales (Garssen et al., 2016). The first challenge, then, is to define measures of spirituality that are not well-being measures dressed up in other words.

A second challenge is to disentangle concepts of spirituality from concepts of religion. The term \enquote{spirituality} was first used to describe the religious practices of ascetics and monks, however, more recently, the term has taken on a broader family of meanings (Swinton, 2001). Spirituality and health research does not converge on a unique set of meanings from the wider universe of popular meanings. Although it is commonplace across most areas of psychological science for researchers to explore different operationalisations of focal concepts, the psychology of spirituality has mostly been conducted with cross-sectional North American samples among participants who affiliate with traditional religion (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky, Ratner, \& Chiu, 2005; Smith et al., 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012). As such, the extent to which the psychology of spirituality and health has repeated or extended the psychology of religion and health remains unclear. Notably, researchers who have operationalised spirituality constructs as distinct from religion constructs have observed a negative association between spirituality and subjective well-being (King et al., 2013). Thus to infer how spirituality affects subjective well-being, it is important to disentangle the concepts of spirituality from concepts of religion, and to carefully investigate features of constructs in populations where these dimensions overlap, partially overlap, and diverge.

A third challenge is to investigate the relationship between spirituality and subjective well-being in nationally diverse samples over individual life-spans. Reliable inference requires observing change within people who differ in dimensions of spirituality over time, and assessing how their lives turn out (Garssen \& Visser, 2016). Previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have relied on cross-sectional samples (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky et al., 2005; Smith et al., 2003; Visser et al., 2010; Yonker et al., 2012). However, To address the key questions in spirituality psychological health research requires longitudinal investigations. Unfortunately the relevant datasets are rare.

@@

Statistical analysis was performed using R version 4.0.2 (2020-06-22). We analyze data from participants who responded to the NZAVS survey at least three four times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of 21,705 (N = 20,979 with complete responses). To model growth in subjective well-being we use generalized linear mixed models. In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator). The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God. To handle the dependencies and heterogeneity introduced from the repeated measures, we include individual ID as an effect modelled as random. We centered and scaled the education, deprivation, and political conservativism. We centred and put into decade-long intervals.
The model equation(s) are described in Equation 1.

@@

Because missing values can result in biased estimates (Blackwell, Honaker, \& King, 2017), we multiply imputed 10 datasets using the Amelia II package in R (Honaker et al., 2011b). We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences. Details of this analysis can be found at: \url{https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing}.

@@

\begin{itemize}
\tightlist
\item
  The effect of Years is positive is not reliable (beta = 2.66e-03, SE = 3.58e-03, std. beta = 3.19e-03, p = 0.458). We infer that Skeptics personal well-being is not increasing in the Skeptic population
\item
  The effect of Beliefs in SpiritExcludesGod is reliably negative (beta = -0.14, SE = 0.03, std. beta = -0.03, p \textless{} .001). We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower personal well-being than skeptics.
\item
  The effect of Beliefs in GodAndSpirit is reliably negative (beta = -0.16, SE = 0.03, std. beta = -0.02, p \textless{} .001). We infer that overall the population that believes in God and a spirit or life force is expected to have a lower Personal Well-being than skeptics.
\item
  The effect of Beliefs in GodExcludesSpirit is not reliably different from the skeptic population (beta = 0.02, SE = 0.05, std. beta = 0.02, p = 0.669).
\item
  The effect of Age.10yrs.C is reliably positive (beta = 0.08, SE = 7.38e-03, std. beta = 0.07, p \textless{} .001). Older people have greater personal well-being.
\item
  The effect of Deprivation (S) is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p \textless{} .001). Deprived people have lower personal well-being.
\item
  The effect of Education (S) is reliably positive (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p \textless{} .001). Educated people greater personal well-being.
\item
  The effect of Employed is reliably positive (beta = 0.09, SE = 0.01, std. beta = 0.05, p \textless{} .001). Employed people have greater personal well-being.
\item
  The effect of identifying as Maori is reliably negative (beta = -0.30, SE = 0.03, std. beta = -0.18, p \textless{} .001).
\item
  The effect of identifying as Pacific is reliably negative (beta = -0.29, SE = 0.05, std. beta = -0.17, p \textless{} .001).
\item
  The effect of identifying as Asian is not reliable (beta = -0.08, SE = 0.05, std. beta = -0.05, p = 0.089).
\item
  The effect of identifying as Male is reliably negative (beta = -0.11, SE = 0.02, std. beta = -0.06, p \textless{} .001).
\item
  The effect of having a Partner is reliably positive (beta = 0.61, SE = 0.02, std. beta = 0.37, p \textless{} .001).
\item
  The effect of Political Conservativism (S) is reliably positive (beta = 0.02, SE = 5.67e-03, std. beta = 0.01, p \textless{} .001).
\item
  The effect of living in an Urban area is reliably negative (beta = -0.02, SE = 0.01, std. beta = -0.01, p \textless{} .05).
\end{itemize}

\hypertarget{personal-well-being-years-x-beliefs}{%
\subsection{Personal Well-Being: Years X Beliefs}\label{personal-well-being-years-x-beliefs}}

@@

\begin{itemize}
\tightlist
\item
  The effect of years \(\times\) belief in a SpiritExcludesGod\ is reliably positive (beta = 0.02, SE = 4.86e-03, std. beta = 0.02, p \textless{} .01). We infer that, unlike the skeptic population, people with beliefs in a spirit or life force who do not believe in a God tend to grow in their personal well-being over time.
\item
  The effect of years \(\times\) in GodAndSpirit is reliably positive (beta = 0.02, SE = 4.41e-03, std. beta = 0.03, p \textless{} .001). We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their personal well-being over time.
\item
  The effect of Years \(\times\) belief in GodExcludesSpirit is not reliable (beta = 1.96e-03, SE = 8.29e-03, std. beta = 2.36e-03, p = 0.813). Thus, God believers who doubt in a Spirit or Life Force start maintain higher personal well-being than do other groups, but they are not expected to grow in personal well-being over time.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  The effect of years is not reliable (beta = -5.50e-04, SE = 2.69e-03, std. beta = -9.32e-04, p = 0.838). We do not infer growth over time in Life Satisfaction among the Skeptic population
\item
  The effect of beliefs in SpiritExcludesGod is reliably negative (beta = -0.08, SE = 0.02, std. beta = 0.01, p \textless{} .001). We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower Life Satisfaction than skeptics.
\item
  The effect of beliefs in GodAndSpirit is not reliable (beta = -0.04, SE = 0.02, std. beta = 0.04, p = 0.105). We do not infer growth over time in Life Satisfaction among this (Skeptic) population.
\item
  The effect of beliefs in GodExcludesSpirit is reliably positive (beta = 0.07, SE = 0.04, std. beta = 0.05, p = 0.062). We infer that overall the population that believes in a God but does not believe in a spirit is not reliably different from the Skeptics population in Life Satisfaction.
\item
  The effect of Age.10yrs.C is reliably positive (beta = 0.09, SE = 5.26e-03, std. beta = 0.10, p \textless{} .001). Older people have greater Life Satisfaction
\item
  The effect of Deprivation (S) is is reliably negative (beta = -0.07, SE = 5.50e-03, std. beta = -0.06, p \textless{} .001). People who are more deprived have lower Life Satisfaction.
\item
  The effect of Education (S) is reliably positive (beta = 0.06, SE = 6.33e-03, std. beta = 0.05, p \textless{} .001). Educated people have greater Life Satisfaction
\item
  The effect of being Employed is reliably positive (beta = 0.08, SE = 9.69e-03, std. beta = 0.07, p \textless{} .001). People who are employed tend to have greater Life Satisfaction
\item
  The effect of identifying as Maori is reliably negative (beta = -0.06, SE = 0.02, std. beta = -0.05, p \textless{} .01). People who belong to the minority Maori group tend to have lower Life Satisfaction.
\item
  The effect of identifying as Pacific is reliably negative (beta = -0.08, SE = 0.04, std. beta = -0.07, p \textless{} .05). People who belong to the minority Pacific peoples group tend to have lower life satisfaction.
\item
  The effect of identifying as Asian is not reliably different from people of European ethnicity in life satisfaction (beta = -0.04, SE = 0.03, std. beta = -0.03, p = 0.223).
\item
  The effect of identifying as Male is reliably negative (beta = -0.19, SE = 0.01, std. beta = -0.16, p \textless{} .001).
\item
  The effect of having a Partner is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p \textless{} .001).
\item
  The effect of Political Conservativism (S) is reliably positive (beta = 0.01, SE = 4.23e-03, std. beta = 0.01, p \textless{} .01).
\item
  The effect of living in an Urban area is reliably negative (beta = -0.02, SE = 8.85e-03, std. beta = -0.02, p \textless{} .05).
\end{itemize}

\hypertarget{life-satisfaction-years-x-beliefs}{%
\subsection{Life Satisfaction: Years X Beliefs}\label{life-satisfaction-years-x-beliefs}}

@@

\begin{itemize}
\tightlist
\item
  The interaction of years \(\times\) beliefs in SpiritExcludesGod is reliably positive (beta = 0.02, SE = 3.66e-03, std. beta = 0.03, p \textless{} .001). We infer that, unlike the Skeptic population, people with beliefs in a Spirit or Life Force who do not believe in a God tend to grow in their life satisfaction over time.

  \begin{itemize}
  \tightlist
  \item
    The interaction of years \(\times\) beliefsGodAndSpirit is reliably positive (beta = 0.01, SE = 3.32e-03, std. beta = 0.02, p \textless{} .001). We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their life satisfaction over time.
  \item
    The effect of years \(\times\) beliefs in GodExcludesSpirit is not reliable (beta = -1.59e-03, SE = 6.25e-03, std. beta = -2.70e-03, p = 0.799). Thus, God believers who doubt in a Spirit or Life Force start maintain higher life-satisfaction than do other groups, but they are not expected to grow in life-satisfaction over time.
  \end{itemize}
\end{itemize}

@@

To adjust for potential bias from missing responses, we conducted a supplementary analysis in which we imputed missing values over the the longitudinal sample. This analysis is describe in Supplement 1. Averaging over the missing responses does not change inference, however we observe growth across all four groups in both personal well-being and life satisfaction. Additionally we observe substantially stronger effects for the slope of Spirit Beliefs on both personal well-being and life-satisfaction. Figures 3 and Figures 4 describe the predicted slopes of change. The inference from this analysis remains the same as from the analysis in which we delete missing values row-wise, by panel wave. In all analyses, those who believe in a spirit or life-force experience greater growth in both personal well-being and life satisfaction over time.

@@

  At the outset of the study, in 2010, the expected subjective well-being as measured using the Personal Well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God (\(\hat{y}\) = 6.50, se =.03) and lower among those who believed in both a Spirit/Life-Force and a God (\(\hat{y}\) = 6.48, se =.03) in comparison both to the skeptical population who neither believed in a God nor a life force (\(\hat{y}\) = 6.64,se =.03) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 6.66, se =.05). (See Figure 2) Similarly, the expected subjective well-being as measured using the Life Satisfaction Inventory is lower among those who believed in Spirit or Life Force but not a God (\(\hat{y}\) = 4.86, se =.02) and lower among those who believed in both a Spirit/Life-Force and a God (\(\hat{y}\) = 4.91, se =.02) in comparison both with the skeptical population who neither believed in a God nor a life force (\(\hat{y}\) = 4.94,se =.02) and the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 5.02, se =.02) (see Figure 3)

@@

  However, over time, belief in a Spirit or Life Force predicts a trajectory of growth in personal well-being. In 2018, personal well-being among those who believe in Spirit or Life Force but not a God had grown to (\(\hat{y}\) = 6.67, se =.03, \(\Delta \hat{y} = 0.17\)). Among those who believed in both a Spirit/Life-Force and a God Personal Well-Being had grown to (\(\hat{y}\) = 6.69, se = 0.02,\(\Delta \hat{y} = .21\)). To place this effect in practical perspective, such growth in personal well-being is about twice as large as the expected benefit of being employed and about a third as large as the expected benefit from having a romantic partner. By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force remained stable across the eight year interval \(\hat{y} = 6.66, se = 0.03, \Delta \hat{y} = .02\)). By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force was constant \(\hat{y} = 6.70, se = 0.04, \Delta \hat{y} = .04\)). Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in life satisfaction. In 2018, life satisfaction among those who believe in Spirit or Life Force but not a God had grown to (\(\hat{y} = 5.01, se = 0.02, \Delta \hat{y} = 0.07\)). In practical terms, such growth is equivalent to the expected gain in life satisfaction from being employed and a sixth as large as the expected benefit to life satisfaction from having a romantic partner. Among those who believed in both a Spirit/Life-Force and a God, life satisfaction had grown to (\(\hat{y} = 5.03, se = 0.02, \Delta \hat{y} = 0.12\)). In practical terms, such growth is equivalent to 1.5 times the expected gain to life satisfaction from being employed and almost a third benefit to life satisfaction from having a romantic partner. By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force remained stable over the eight year interval \(\hat{y} = 4.94, se =0.02, \Delta \hat{y} = 0\)). Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force was constant \(\hat{y} = 5.00, se = 0.03, \Delta \hat{y} = -.02\)) (See Figure 3).
